Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire

### 2016-10-05 manuscript/manuscript.Rnw
Updates to manuscript

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, it is unknown whether increases in woody species and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that warming doubled tall shrub seedling growth rates and could increase survival. Warming had limited or no effect on shrub recruitment, post-fire gap-infilling by grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival.
\end{abstract}

@@

Field manipulative experiments have been widely used to investigate the effects of climate change on plant communities. These experiments have provided invaluable information on how climate change directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}. However, because disturbances vary both spatially and temporally, most have focused on mature plant responses in relatively undisturbed vegetation \citep{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}. Consequently, little is known about how climate change influences seedling demographic rates \citep{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions.

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ua, Batllori:2009fe, Frost:2013tm}, that climatic warming is likely to increase woody species growth rates \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp}, and that shrubs are the most flammable component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. The combination of this evidence indicates that more frequent or severe fire may increase recruitment opportunities for woody species, which when coupled with higher growth and survival rates caused by rising temperatures, will exacerbate shrub expansion \citep{Racine:2004tn,Lantz:2013fh} and ultimately increase the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. In effect, warming could strengthen an existing feedback loop that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback loop. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid blue lines = known mechansims; Dashed black lines = mechansims that may exacerbate or diminish this feedback loop but which we have a paucity of information on. (b) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dotted line) that could break this feedback.}
\label{fig1}
\end{figure}

@@

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To determine drivers of shrub establishment and how warmer post-fire conditions affect shrub cover, we combined field observations with a warming experiment that used seedlings of four Australian evergreen obligate seeding shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub). This combination of data sources allowed us to examine: 1) landscape scale drivers of shrub seedling establishment; 2) how warmer post-fire conditions influence rates of grass recovery and shrub seedling recruitment, growth and mortality; 3) how proximity to grass affect shrub seedling demographic rates; and 4) whether such grass effects change under warmer post-fire conditions.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5 m$^2$) \textit{Grevillea australis} shrubs in open heathland to create patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance in open heathland burnt by wildfire. After creating the bare ground patches we collected seedlings of dominant alpine shrub species from a nearby (\textless 2 km) site of similar altitude burnt by a late 2006 wildfire. We collected shrub seedlings of two dominant open heathland species \textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland \textit{Phebalium squamulosum} (Rutaceae; a tall shrub). All four species are killed by fire \citep{Camac:2013eb}.

@@

\section*{Results}
\begin{figure}[ht]
\centering
\includegraphics{figures/fig2.pdf}
\caption{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (twig diameter), altitude, Topographic Wetness Index (TWI) and adult density. (a) Centered and standardized model coefficients (on the log scale); and effects of: (b) burning, (c) fire severity, (d) altitude, (e) topographic wetness and (F) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{figure}

\begin{figure*}[!h]
\centering
\includegraphics{figures/fig3.pdf}
\caption{Observed and projected growth trajectories of four dominant Australian alpine shrubs. (a) Observed height growth: Thick lines with error bars represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (b) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red and orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively). Vertical lines delimit year in which maximum height is obtained}
\label{fig3}
\end{figure*}

\subsection*{Drivers of shrub seedling establishment}
We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by post-fire twig diameters\textemdash see Supporting Information) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Fig. \ref{figs1}) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with larger post-fire twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range \citep[equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature;][]{Brown:1989vx}. The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Fig. \ref{figs2}) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no strong effect of Topographic Wetness Index for either species.

@@

\subsection*{Experimental warming on shrub seedling mortality}
\begin{figure}[!h]
\centering
\includegraphics{figures/fig4.pdf}
\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) annual probability of death for each species growing in control (blue) and warmed (red) conditions.}
\label{fig4}
\end{figure}

@@

@@

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems \citep{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates conditions that allow shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates and potentially increase their survival. These demographic effects will ultimately result in shrub thickening and expansion into grasslands. Coupled with field \citep{Williams:2006tl}, experimental \citep{Fraser:NHT5Kx_W} and paleoecological \citep{Higuera:2008dw} evidence, which indicate that shrubs are the most flammable component of alpine and tundra ecosystems, our results suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine and arctic landscapes under a warmer environment. This further strengthens the feedback \citep{Higuera:2008dw, Higuera:2009ga} by potentially increasing the frequency or severity of fires, which then creates further recruitment opportunities for shrubs with little demographic impact of neighbouring tussock grass.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent obligate seeding shrubs reaching reproductive age and thereby exhausting the seedbank. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including in Australia, where current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow obligate seeding species to reach reproductive maturity sooner, and consequently, may increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq} or herbivory \citep{Post:2008wz} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways \citep{MyersSmith:2011fp}.

Our findings provide mechanistic understanding as to why shrub cover has increased, often at the expense of grasslands, in many alpine and arctic ecosystems \citep{Racine:2004tn,MyersSmith:2011fp,Frost:2013tm,Lantz:2013fh}. But more importantly, our results provide evidence for underlying processes that could result in a warming-fire-shrub feedback that has been hypothesized in paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga}. Based on current observations and predictions, average global temperature has already increased by 0.85\textdegree C since 1880 and is expected to rise by as much as 4.8\textdegree C by 2100 \citep{IPCC:2013vv}. In alpine and tundra environments, temperatures \citep{ChapinIII:2005js}, shrub cover \citep{MyersSmith:2011fp} and the frequency and severity of fire \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp} have all increased in the last few decades. These changes mean that the warming-shrub-fire feedback loop identified here has already strengthened, which could cause grasslands and other non-woody communities to transition to an alternative state with more shrubs and more fire, both of which are likely to have consequences for carbon sequestration, water supply and biodiversity.

>>>>

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, it is unknown whether increases in woody species and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen fire-killed alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that warming doubled tall shrub seedling growth rates and could increase survival. Warming had limited or no effect on shrub recruitment, post-fire gap-infilling by grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival.
\end{abstract}

@@

Field manipulative experiments have been widely used to investigate the effects of climate change in alpine and tundra plant communities. These experiments have provided invaluable information on how climate directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}. However, because the general perception is that alpine and tundra vegetation is predominately clonal \citep{Billings:1968bn}, despite contradictory evidence \citep[e.g.][]{Venn:2009bw,Briceno:2015iv}, coupled with the fact that disturbances vary both spatially and temporally, most studies have focused on mature plant responses in relatively undisturbed vegetation \citep{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}. Consequently, little is known about how climate change influences seedling demographic rates \citep{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions.

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ua, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover \citep[e.g. undisturbed alpine tussock grassland;][]{Williams:1987ua}. Evidence also indicates that shrubs can have strong supressive effects on other life forms \citep{MyersSmith:2011fp}, that climatic warming is likely to increase woody species growth rates \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp}, and that shrubs are the most flammable component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. The combination of this evidence indicates that more frequent or severe fire may increase recruitment opportunities for woody species, which when coupled with higher growth and survival rates caused by rising temperatures, will exacerbate shrub expansion \citep{Racine:2004tn,Lantz:2013fh} and ultimately increase the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. In effect, warming could strengthen an existing feedback loop that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback loop. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid lines = known mechansims; Dashed lines = mechansims that may exacerbate or diminish this feedback loop but which we have a paucity of information on. (b) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dashed line) that could break this feedback. Blue = positive relationships, Red = negative relationships.}
\label{fig1}
\end{figure}

@@

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To determine drivers of shrub establishment and how warmer post-fire conditions affect shrub vital rates, we combined field observations with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub). This combination of data sources allowed us to examine: 1) landscape scale drivers of shrub seedling establishment; 2) how warmer post-fire conditions influence rates of grass recovery and shrub seedling recruitment, growth and mortality; 3) how proximity to grass affect shrub seedling demographic rates; and 4) whether such grass effects change under warmer post-fire conditions.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5 m$^2$) \textit{Grevillea australis} shrubs in open heathland to create patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance in open heathland burnt by wildfire. After creating the bare ground patches we collected seedlings of dominant alpine shrub species from a nearby (\textless 2 km) site of similar altitude burnt by a late 2006 wildfire. We collected shrub seedlings of two dominant open heathland species \textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland \textit{Phebalium squamulosum} (Rutaceae; a tall shrub). We focused on these four species because they are all fire-killed \citep[the dominant shrub post-fire strategy in the Australian mainland Alps;][]{Walsh:2004vk}; common in the Australian Alps; and under climatic warming, are likely candidates to invading non-shrubby plant communities such as alpine grasslands and herbfields.

@@

\begin{figure}[!htb]
\centering
\includegraphics{figures/fig2.pdf}
\caption{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (twig diameter), altitude, Topographic Wetness Index (TWI) and adult density. (a) Centered and standardized model coefficients (on the log scale); and effects of: (b) burning, (c) fire severity, (d) altitude, (e) topographic wetness and (F) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{figure}

\begin{figure*}[!htb]
\centering
\includegraphics{figures/fig3.pdf}
\caption{Observed and projected growth trajectories of four dominant Australian alpine shrubs. (a) Observed height growth: Thick lines with error bars represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (b) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red and orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively). Vertical lines delimit year in which maximum height is obtained}
\label{fig3}
\end{figure*}

\section*{Results}
\subsection*{Drivers of shrub seedling establishment}

We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by post-fire twig diameters\textemdash see Supporting Information) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Fig. \ref{figs1}) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with larger post-fire twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range \citep[equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature;][]{Brown:1989vx}. The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Fig. \ref{figs2}) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no strong effect of Topographic Wetness Index for either species.

@@

\subsection*{Experimental warming on shrub seedling mortality}

@@

\begin{figure}[!h]
\centering
\includegraphics{figures/fig4.pdf}
\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) annual probability of death for each species growing in control (blue) and warmed (red) conditions.}
\label{fig4}
\end{figure}

@@

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems \citep{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates bare ground required for shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates and potentially increase their survival. These demographic effects will ultimately manifest in shrub thickening within and at heathland boundaries, which in turn, will facilitate shrub establishment into shrub-free communities by decreasing dispersal distances. Coupled with slow tussock grass gap infilling and, field \citep{Williams:2006tl}, experimental \citep{Fraser:NHT5Kx_W} and paleoecological \citep{Higuera:2008dw} evidence which indicate that shrubs are the most flammable component of alpine and tundra ecosystems  \citep[particularly moreso then tussock grassland; ][]{Fraser:NHT5Kx_W}, our results suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine and arctic landscapes under a warmer environment. This further strengthens the feedback \citep{Higuera:2008dw, Higuera:2009ga} by increasing the frequency or severity of fires, which then creates more shrub recruitment opportunities both within and beyond heathland boundaries (i.e. via more fire-induced tussock death or increased exposure of inter-tussock gaps) with little demographic impact of neighbouring tussock grass.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent fire-killed shrubs reaching reproductive age and thereby exhausting the seedbank. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including in Australia, where current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow fire-killed species to reach reproductive maturity sooner, and consequently, may increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq} or herbivory \citep{Post:2008wz} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways \citep{MyersSmith:2011fp}.

By focusing on seedlings, the life stage most vulnerable to climate and disturbance, and which determines the long-term persistence of a species and its capacity to establish in new areas \citep{Walck:2011dm, Briceno:2015iv}, our findings provide mechanistic understanding as to why shrub cover has increased, often at the expense of grasslands, in many alpine and arctic ecosystems \citep{Racine:2004tn,MyersSmith:2011fp,Frost:2013tm,Lantz:2013fh}. But more importantly, our results provide evidence for underlying processes that could result in a warming-fire-shrub feedback that has been hypothesized in paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga}. Based on current observations and predictions, average global temperature has already increased by 0.85\textdegree C since 1880 and is expected to rise by as much as 4.8\textdegree C by 2100 \citep{IPCC:2013vv}. In alpine and tundra environments, temperatures \citep{ChapinIII:2005js}, shrub cover \citep{MyersSmith:2011fp} and the frequency and severity of fire \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp} have all increased in the last few decades. These changes mean that the warming-shrub-fire feedback loop identified here has already strengthened, which could cause grasslands and other non-woody communities to transition to an alternative state with more shrubs and more fire, both of which are likely to have consequences for carbon sequestration, water supply and biodiversity.

### 2016-10-05 manuscript/supplementary.Rnw
Updates to manuscript

We collected seedlings from a site burnt by a late 2006 fire because it allowed us to collect sufficient numbers of seedlings from species occurring in both open and closed heathland. We focused on \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) because they are all: obligate seeders post-fire (i.e. fire killed); common in the Australian Alps; and under climatic warming, are likely to invading non-shrubby plant communities such as alpine grasslands and herbfields. For each species, we collected seedlings with a range of heights (\Sexpr{range_init_size(complete_otc_seedling_data)[1]} to \Sexpr{range_init_size(complete_otc_seedling_data)[2]} cm) using a forestry planter (12.5 cm deep by 5.5 cm wide). The size of the planter was selected to minimize root damage across the range of seedling sizes and to allow for the extraction of a constant volume (378 cm$^3$) of intact soil and roots. The planter also minimized soil disturbance during seedling extraction and replanting.

Four seedlings of each species were then transplanted into 70 cm$^2$ of burnt bare ground in the center of each experimental plot. An additional four \textit{Grevillea} seedlings were transplanted into naturally occuring inter-tussock gaps of varying size surrounding this bare ground. Only \textit{Grevillea} seedlings were transplanted into inter-tussock gaps because of space constraints but also because it is the dominant open heathland shrub species (i.e. occurs closest to the shrub/grassland boundary) and is therefore the most likely to invade grasslands. The combination of bare ground and inter-tussock seedlings meant that each experimental plot had a total of 20 seedlings per 1 m$^2$ (Fig. \ref{figs8}). However, the two sets of seedlings (i.e. bare ground vs inter-tussock) were treated separately because they were spatially confounded (e.g. center vs edge of the 1 m$^2$ opening) and addressed different questions. Seedlings in bare ground patches (i.e. the center of the plot) were used to assess whether warmer post-fire conditions altered growth and mortality rates in four dominant alpine shrub species. By comparison, \textit{Grevillea} seedlings growing in inter-tussock gaps were used to determine whether inter-tussock gap size altered growth and mortality rates, and whether such effects were modified by warmer conditions.

>>>>

We collected seedlings from a site burnt by a late 2006 fire because it allowed us to collect sufficient numbers of seedlings from species occurring in both open and closed heathland. We focused on \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) because they are all fire-killed (the dominant shrub post-fire strategy in the Australian mainland Alps; \citet{Walsh:2004vk}); common in the Australian Alps; and under climatic warming, are likely candidates to invading non-shrubby plant communities such as alpine grasslands and herbfields. Resprouter shrub species were not examined in this study because we were unable to locate sufficient numbers of seedlings; and are unlikely to be the primary contributors to shrub expansion in these landscapes. For each species, we collected seedlings with a range of heights (\Sexpr{range_init_size(complete_otc_seedling_data)[1]} to \Sexpr{range_init_size(complete_otc_seedling_data)[2]} cm) using a forestry planter (12.5 cm deep by 5.5 cm wide). The size of the planter was selected to minimize root damage across the range of seedling sizes and to allow for the extraction of a constant volume (378 cm$^3$) of intact soil and roots. The planter also minimized soil disturbance during seedling extraction and replanting.

Four seedlings of each species were then transplanted into 70 cm$^2$ of burnt bare ground in the center of each experimental plot. An additional four \textit{Grevillea} seedlings were transplanted into occuring post-fire inter-tussock gaps of varying size surrounding this bare ground. Only \textit{Grevillea} seedlings were transplanted into inter-tussock gaps because of space constraints but also because it is the dominant open heathland shrub species (i.e. occurs closest to the shrub/grassland boundary) and is therefore the most likely to invade grasslands. The combination of bare ground and inter-tussock seedlings meant that each experimental plot had a total of 20 seedlings per 1 m$^2$ (Fig. \ref{figs8}). However, the two sets of seedlings (i.e. bare ground vs inter-tussock) were treated separately because they were spatially confounded (e.g. center vs edge of the 1 m$^2$ opening) and addressed different questions. Seedlings in bare ground patches (i.e. the center of the plot) were used to assess whether warmer post-fire conditions altered growth and mortality rates in four dominant alpine shrub species. By comparison, \textit{Grevillea} seedlings growing in inter-tussock gaps were used to determine whether inter-tussock gap size altered growth and mortality rates, and whether such effects were modified by warmer conditions.
